---
title: "wine_kmeans"
author: "Jerry Xu"
date: "3 August 2018"
output: pdf_document
---

```{r}
library(dplyr)
library(tidyr)
library(factoextra)
library(kmed)
```
Helpful references:

https://www3.nd.edu/~steve/computing_with_data/24_dplyr/dplyr.html



### Gower distances:

1) Manhatten distance for __continous variables__
2) __Ordinal variables__ are ranked then distance calculated with Manhatten
3) __Binary variables__ <- Jaccard matching coefficient is used. (when both variables lack attribute, 
    JMC treats it as either match or mismatch)
4) Dice matching coeffient with __nominal variables__

```{r}

header = c("Class" ,"Alcohol" , "Malic acid", "Ash", "Alcalinity of ash", "Magnesium", "Total phenols", 
          "Flavanoids", "Nonflavanoid phenols", "Proanthocyanins", "Color intensity", "Hue", "OD280/OD315 of diluted wines", "Proline")

wine_truth = read.csv("wine.data", col.names=header)

wine_truth$Class <- as.factor(wine_truth$Class)
wine_truth
```

### K-means clustering

Used for numeric only, once distances are computed for qualitative variables.

```{r}

wine_test = select(wine_truth, -Class)
wine_test

```

```{r}
wine_groups = kmeans(wine_test, 3, iter.max=10, nstart=10)
```

```{r}
wine_groups$size
as.vector(count(wine_truth, var=Class))$n
```

```{r}
fviz_nbclust(wine_test, kmeans, method="wss") + geom_vline(xintercept=4, linetype=2)
# wss = within sum of sqs 
```



```{r}
wine_groups = kmeans(wine_test, 3, iter.max=10, nstart=8)
fviz_cluster(wine_groups, data=wine_test, 
             ellipse.type = "norm", 
             ggtheme=theme_minimal(),
             star.plot= TRUE)
```



```{r}

wine_test_scale <- scale(wine_test)
wine_groups_scale = kmeans(wine_test_scale, 3, iter.max=10, nstart=10)

fviz_cluster(wine_groups_scale, data=wine_test_scale, 
             ellipse.type = "norm", 
             ggtheme=theme_minimal()
             )
```

```{r}
wine_test
```


```{r}
wine_test_scale2 <- scale(wine_test)
print(dim(wine_test_scale2))
wine_test_dist2 <- dist(wine_test_scale2)


# What does the dist() do?
# wine_test_dist2
length(wine_test_dist2)
wine_test_dist2 # printout is omitted
```


```{r}
wine_groups_scale2 = kmeans(wine_test_dist2, 3, iter.max=10, nstart=10)

fviz_cluster(wine_groups_scale2, data=wine_test_scale2, 
             ellipse.type = "norm", 
             ggtheme=theme_minimal()
             )
```

If there were quantitative variables, then we can use Gower's distances as above.

^ How do we deal with this?

First let's simulate some fake categorical variables

```{r}
a <- matrix(sample(1:4, 7*3, replace = TRUE), 7, 3)
matching(a, a)
a

b <- matrix(sample(1:2, 7*3, replace = TRUE), 7, 3)
b
coocurance(b)
```


```{r}

```

